---
title: "TEMPLATE - alternative chart types practice"
author: "Your Name"
date: 2026-03-04
toc: true
---

::: {.callout-note}
This template accompanies the [alternative chart types practice slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-alt-charts.html#/title-slide){target="_blank}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

# Waffle Chart (Sasquatch sightings)

- Typefaces used: Ultra & Josefin Sans
- Colors used: "#357266", "#FFB813", "#983A06", "#005F71", "#757473", "#EFEFEF"

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-align: "center"
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(waffle)
library(showtext)

#..........................import data...........................
bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-13/bigfoot.csv')

bigfoot_ca <- bigfoot %>% 
  filter(state == "California", !is.na(season), season != "Unknown") %>% 
  count(season) %>% 
  mutate(season = factor(season, levels = c("Spring", "Summer", "Fall", "Winter"))) %>% 
  
  # forces the order!
  arrange(season)

 # deframe() # deframing required for waffle


# color palette
seasons <- c("Spring" = "#357266", 
             "Summer" = "#FFB813",
             "Fall" = "#983B07",
             "Winter" = "#005F71")

ggplot(bigfoot_ca, aes(fill = season, values = n)) +
  geom_waffle(n_rows = 10, make_proportional = TRUE,
              color = "white") +
  coord_equal() +
  theme_void() +
  labs(title = "Bigfoot Sightings in California by Season", fill = "Season") +
  scale_discrete_manual(values = seasons, aesthetics = "fill")
```

# Bump chart (highest paying occupations)

- Typefaces used: Passion One & Oxygen
- Colors used: `viridisLite::magma()` & "#2A114E", "#6D6B71", "#FFF8F4"

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggbump)
library(ggtext)
library(showtext)

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
```

# Stream graph (energy production)

- Typefaces used: Exo 2 & Open Sans
- Colors used: `paletteer::paletteer_d("MetBrewer::Archambault", n = 7)`

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggstream)
library(ggtext)
library(showtext)

#..........................import data...........................
capacity <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-05-03/capacity.csv")
```











