---
title: "TEMPLATE - colors"
author: "Your Name"
date: 2026-02-04
toc: true
---

::: {.callout-note}
This template follows [lecture 5.2 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture5.2-colors-slides.html#/title-slide){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Class Notes

Some places to start/things to consider:
- is using color the best and/or only way to visually represent your variables?
- emotions?
- branding?
- audience?
- data types
- how is this data/these groups commonly represented?
- consider meanings of color in different cultures

Color scales:
- quantitative
  - sequential and unclassed
  - sequential and classed (specific colors picked for specific bins)
  - diverging and unclassed
  - diverging and classed // diverging typically used for "over an average or under an average", or to represent different sides
  
Categorical:
- no more than 7 hues 
- hues assigned to each bin should be distinguishable 

Sequential:
- darker colors = higher numbers (except for elevation!)

Diverging:
- central value is assigned a light color (light gray is best)

Accessibility:
- rainbow color map --> no natural perceived ordering (higher or lower)
- rainbow color map are not good for color blindness!
- perhaps use an improved rainbow colormap! but sitll not good

- keep colors consistent across viz

## Setup

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(palmerpenguins)
library(tidyverse)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                              create base plots                           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# requires categorical color scale ----
cat_color_plot <- ggplot(na.omit(penguins), 
                         aes(x = bill_length_mm, y = bill_depth_mm, 
                             color = species, shape = species)) +
  geom_point(size = 4, alpha = 0.8)

cat_color_plot

# requires continuous color scale ----
con_color_plot <- ggplot(na.omit(penguins), 
                         aes(x = bill_length_mm, y = bill_depth_mm, 
                             color = body_mass_g)) +
  geom_point(size = 4, alpha = 0.8) 

con_color_plot
```

## Colors for inclusive & accessible design

### Viridis scales

- Check out the [documentation](https://ggplot2.tidyverse.org/reference/scale_viridis.html) for additional viridis palette options. 

```{r}
#| eval: true
#| echo: true
#| warning: false

# discrete viridis scales ----
cat_color_plot + 
  scale_color_viridis_d(option = "viridis") # d for discrete variables

# continuous color scales ----
con_color_plot +
  scale_color_viridis_c(option = "magma") # c for continuous 
```

### RColorBrewer scales

- Exploring available palettes and generating HEX codes (or check out the [web-based interface](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)):

```{r}
#| eval: true
#| echo: true
# display only colorblind-friendly RColorBrewer palettes ----
RColorBrewer::display.brewer.all(colorblindFriendly = TRUE)

# preview palette with your number of desired colors ----
RColorBrewer::display.brewer.pal(n = 3, name = "Dark2")

# print the HEX codes of your palette ----
RColorBrewer::brewer.pal(n = 3, name = "Dark2")

# or save hex codes as a vector ----
my_brewer_pal <- RColorBrewer::brewer.pal(n = 3, name = "Dark2")
```

- Using RColorBrewer palettes

```{r}
#| eval: true
#| echo: true
#| warning: false

# for qualitative palettes ----
cat_color_plot +
  scale_color_brewer(palette = "Dark2") # will automatically select first 3 colors for you

# for unclassed continuous color scales ----
con_color_plot +
  scale_color_distiller(palette = "BuPu", direction = 1) # direction = 1 reverses order of color scale

# for classed continuous color scales ----
con_color_plot +
  scale_color_fermenter(palette = "YlGnBu")



# Accessibility tip:
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, fill = body_mass_g)) + # change to fill object introduces border around points 
  geom_point(shape = 21, size = 4, alpha = 0.8) +
  scale_fill_fermenter(palette = "YlGnBu")
```


## `{paletteer}`

A few different ways to use `{paletteer}` palettes, but my preferred approach takes two steps: 

1. create vector of colors using `paletteer_*()`
2. apply your palette using the appropriate `ggplot::scale_*()` function


```{r}
#| eval: true
#| echo: true
# import {paletteer} ----
library(paletteer)
```


```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false


# Discrete Data ------

# Step 1: Create palette
pal_d <- paletteer_d("wesanderson::GrandBudapest1", n = 3) # if run console, will return hex codes with colors!

#Step 2 Apply palette using `color` variant 
cat_color_plot +
  scale_color_manual(values = pal_d)

#Step 2 Apply palette using `fill` variant)
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_histogram() +
  scale_fill_manual(values = pal_d)
```



```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# Continuous Data ------

## Step 1:  create palette
pal_c <- paletteer_c("scico::batlow", n = 5, direction = -1)

#Step 2: Apply palette using `gradientn` variant
con_color_plot + 
  scale_color_gradientn(colors = pal_c) # gradient, when continuous color; n because we have a particular number of colors

#Step 2: Apply palette using `stepsn` variant
con_color_plot + 
  scale_color_stepsn(colors = pal_c) # discrete 
```

## Some final palette tips

### Save palette outside of plot

```{r}
#| eval: true
#| echo: true
#| warning: false
# create palette ----
my_palette <- c("#32DE8A", "#E36414", "#0F4C5C")

# plot with custom palette ----
cat_color_plot +
  scale_color_manual(values = my_brewer_pal)



```

### Set color names

```{r}
#| eval: true
#| echo: true
#| warning: false
# create palette ----
my_palette_named <- c("Adelie" = "#32DE8A", "Chinstrap" = "#E36414", "Gentoo" = "#0F4C5C") # ensures that colors stay consistent across viz!

# apply to plot (all penguins) ----
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_palette_named)

# apply to plot (just adelie & gentoo) ----
penguins |> 
  filter(species != "Chinstrap") |> 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = my_palette_named)
```

